---
title: "Codyn Tutorial"
author: "Jory Griffith"
date: "4/7/2022"
output: 
  html_document:
    toc: true
    theme: united
    toc_float: true
---
# Load Packages
We will be using the `codyn` package for this tutorial. We will also use `ggplot2` and `dplyr` for plotting and cleaning.
```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(codyn)
library(dplyr)
```


# Load Data
These data are from the Turkey Lakes watershed experiment. This dataset contains the abundances of benthic invertebrate species measured in the May and June from 1995 to 2001. They recorded the type and number of invertebrates in an area of the stream, with 10 replicates per sampling event. These forests next to these streams were logged in 1997 with varying harvest intensities, with low, medium, and high intensity harvest and a control stream with no forest harvest. We will using the `codyn` package to analyze the diversity of these benthic invertebrates over time. 
```{r}
data<-read.csv("stream_invertebrates.csv")
head(data)
```

For many functions, the `codyn` package can only look at data from one stream at a time. So first we will separate each stream into it's own data frame.
```{r, message=FALSE, warning=FALSE}
# high intensity logging treatment
subset_codyn_high <- data %>%
  filter(treatment == "high")
# medium intensity logging treatment
subset_codyn_med <- data %>%
  filter(treatment == "medium")
# low intensity logging treatment
subset_codyn_low <- data %>%
  filter(treatment == "low") 
# control stream (no logging)
subset_codyn_ctrl <- data %>%
  filter(treatment=="control")
```


# Using the `codyn` package to analyze diversity data over time
Ecologists are often interested in how diversity changes over time in one place. This is particularly interesting when you are looking at how species diversity responds to environmental change. This environmental change can be a single disturbance event such as a forest fire or more long-term change such as temperature increases due to climate change. The Turkey Lakes dataset provides us with an opportunity to look at how the diversity of benthic invertebrates changes over time and in response to logging. We can compare the diversity of different streams with different logging intensities and we can also compare the diversity of streams.

The `codyn` package allows us to analyze the change in biodiversity over time using different metrics.

# Using `community_structure()`
`community_structure()` is a function in the package `codyn` that calculates species richness and evenness. 

## Species Richness and Evenness
For example, we can use `community_structure()` to look at the change in species richness and evenness over time in one of the Turkey Lake streams. We will choose the stream with high logging intensity as an example. This function can only look at one stream at a time so make sure to subset out the stream that you are interested in.

Let's take a look at the high intensity logging stream data.
```{r}
head(subset_codyn_high)
```

As you can see, there is a column for year, species, replicate, and abundance.

The community structure function has five arguments. The first is `df` where you put in the data frame of interest, in our case `subset_codyn_high`. The `time.var` argument is where you put in your time variable (`year`), the abundance.var argument is where you put the column that holds the abundance variable (`abundance`) and the `replicate.var` argument holds the column with the replicates (`replicate`). In the `metric` argument, you can specify the metric of evenness that you want the function to calculate. There are three possible metrics: `Evar`, `SimpsonEvenness`, and `EQ`, which all calculate evenness in different ways. We will use Simpson's Evenness in this tutorial. Remember, if you ever need a refresher on the syntax you can type `?community_structure` into the R script.
```{r}
richness_high<-community_structure(df=subset_codyn_high, time.var="year", abundance.var="abundance", replicate.var="replicate", metric="SimpsonEvenness")
```

We can take a look at the output.
```{r}
head(richness_high)
```

As you can see, this returns a data frame with the year, replicate, richness calculation, and evenness measure. It calculates richness and evenness separately for each replicate sample in the stream, so we have a measure of variability within samples as well as variability across years. 

### Plot species richness

We can plot this data frame to see the changes over time.
```{r}
ggplot()+
  geom_point(richness_high, mapping=aes(x=year, y=richness))+
  theme_classic()+
  labs(y="Species Richness", x="Year")+
  geom_vline(xintercept=1997, linetype=2, color="red")
```

Here is a plot of species richness, with a different value for each replicate. The red line is the year that the streams were logged. Do you see any change? Do you think there are any significant patterns? How does the variability between replicates compare to the variability between years?

### Plot species evenness
We can also plot species evenness over time.
```{r}
ggplot()+
  geom_point(richness_high, mapping=aes(x=year, y=SimpsonEvenness))+
  theme_classic()+
  labs(y="Simpson's Evenness", x="Year")+
  geom_vline(xintercept=1997, linetype=2, color="red")
```

Do you see any patterns in the evenness between years?

# Using `community_diversity()`

## Shannon's diversity
Species richness and evenness are very simple ways to look at species diversity. However, as you learned from the rank abundance curves, species have different abundances within the community, with some species having very high abundance and most species having low abundance. Species richness does not take into account the relative abundances of each species, which limits the inferences we can make. The Shannon-Weiner diversity metric does take into account these relative abundances when calculating the diversity of a community and allows us to make more meaningful inferences about how biodiversity is changing over time.

The `codyn` package allows you to calculate the Shannon diversity index for a single site over time using the `community_diversity()` function. We will again use the data from the stream with the high logging intensity. Luckily, this function has the exact same syntax as the `community_structure()` function! The only difference is that the options for the `metric` argument are `Shannon` and `InverseSimpson`. We will first use `Shannon`.

```{r}
shannon_div_high<-community_diversity(df=subset_codyn_high, time.var="year", abundance.var="abundance", replicate.var="replicate", metric="Shannon")
```

Let's take a look at the output for this function.
```{r}
head(shannon_div_high)
```

You can see it gives us a data frame of `year`, `replicate`, and `Shannon`, which is our measure of Shannon diversity.

### Plot Shannon's diversity
We can plot these to see change over time.
```{r}
ggplot()+
  geom_point(shannon_div_high, mapping=aes(x=year, y=Shannon))+
  theme_classic()+
  labs(y="Shannon Diversity", x="Year")+
  geom_vline(xintercept=1997, linetype=2, color="red")
```

## Simpson's diversity
We can also use the same function to calculate Simpson's diversity by simply changing the `metric` argument to `InverseSimpson`. 
```{r}
simpson_div_high<-community_diversity(df=subset_codyn_high, time.var="year", abundance.var="abundance", replicate.var="replicate", metric="InverseSimpson")
```

We can look at the output.
```{r}
head(simpson_div_high)
```

### Plot Simpson's diversity
And then plot.
```{r}
ggplot()+
  geom_point(simpson_div_high, mapping=aes(x=year, y=InverseSimpson))+
  theme_classic()+
  labs(y="Simpson Diversity", x="Year")+
  geom_vline(xintercept=1997, linetype=2, color="red")
```

How do these patterns compare to the patterns of Shannon's diversity over time? How do the values compare?

# Activity
Calculate richness, evenness, and diversity using both the Shannon and Simpson indices for the streams with low intensity logging, high intensity logging, and no logging. The data for these streams are `subset_codyn_low`, `subset_codyn_high`, and `subset_codyn_ctrl`. Plot these indices. Do you see any patterns across years? How do these compare to the patterns in the high intensity logging stream?
























