---
title: "Tukey Lake Bentic Invertebrates"
author: "Jory Griffith"
date: "4/5/2022"
output: html_document
---

```{r}
library(ggplot2)
library(tidyverse)
```


# Leaf litter in traps
Metadata says:
"Blow-in litter traps measured 0.46 x 0.76 meters but area assumes the long dimension squared (0.5776 meter squared) and so are equivalent to the amount deposited in a normal litter trap if the side wall was down."

These were placed beside the stream for a number of days that is specified in the daysout_number.

The website says that the stream catchments that were harvested were c31, c33, and c34. These were all harvested at different intensities in 1997.

Non-harvested reference catchments were 32; 37; 38; 39; 42; 60; 96; 97; 98; 99 (99U is above the road and 99L is below the road. 99UU and 99UL are respectively more upstream of 99U); 100; 101; 102; NC (Norburg Creek; NCU is upper and NCL is lower).

Trees were harvested in the summer of 1997. Riparian buffers were waived for experimental purposes; prescription was to selectively harvest right to stream edge although no machine traffic was allowed within 3 meters of the stream channels. Catchment 31 harvested at high intensity diameter limit cut (all trees greater than 20 centimeteres were felled then delimbed and removed; all trees between 10-20 centimeter diameter were felled) (average 89% basal area removal); catchment 33 harvested at low intensity selection cut (average 29% basal area removal); catchment 34U harvested at medium intensity shelterwood cut (average 42% removal); catchment 34L harvested at high intensity diameter limit cut (average 89% basal area removal) (34U 34M 34L are in the same catchment; 34L is below the road down to Norburg; 34M and 34L are contiguous above the road; 34M is from the road to 310 meters upstream and then 34U begins at 320 meters).

### Load Data
```{r}
blow_litter<-read.csv("TLW_litterBlowInTraps.csv")
head(blow_litter)
```

```{r}
str(blow_litter)
```

I want to see how many years of data there is and how many catchments.
```{r}
unique(blow_litter$catchment)
unique(blow_litter$daysout_number)
unique(blow_litter$year)
```

I am not sure what the L and M numbers are but stream 34 was harvested and 97 was not so I can start by comparing those.

It looks like the number of days that the leaf litter was left out was always 21 days so there shouldn't be any confounding effects of the time that the trap was left out.

This data was only taken in one year, 2000, which is 3 years post-logging. The sampling effort was from May to October of that year.

There were 10 replicate traps per sampling effort.

## Plot 1
First I will just plot the total litter by e
```{r}
ggplot(blow_litter)+
  geom_col(aes(x=catchment, y=dryleaf_g, fill=date_in))
```

Ok, so essentially there is more litter in the unharvested catchment and there is more litter in the fall (unsuprising). 

Catchment 34L was harvested at high intensity, catchment 34M is the same catchment but above the harvested section.

catchment 34U harvested at medium intensity shelterwood cut (average 42% removal); catchment 34L harvested at high intensity diameter limit cut (average 89% basal area removal) (34U 34M 34L are in the same catchment; 34L is below the road down to Norburg; 34M and 34L are contiguous above the road; 34M is from the road to 310 meters upstream and then 34U begins at 320 meters).

## Plot 2: Organic leaf
```{r}
ggplot(blow_litter)+
  geom_col(aes(x=catchment, y=organicleaf_gperm2, fill=date_in))
```
Pretty much the same situation.


# Leaf Disc Decomposition
This is what the metadata says for this dataset:
"Leaf disks were placed in streams to measure decomposition by microbes. Ten disks were placed in a 500 micron mesh bag and anchored in the stream for a single replicate. The position of leaf disks in the stream were either anchored in the stream (surface) or anchored and buried 1 centimeter below the surface of the streambed (buried). 5 to 8 replicates were placed per catchment."

## Load data
```{r}
leaf_disc<-read.csv("TLW_leafdiskDecomposition.csv")
head(leaf_disc)
str(leaf_disc)
```

Looking at the number of unique catchments, years, etc.
```{r}
unique(leaf_disc$catchment)
# 17 unique catchments
unique(leaf_disc$year)
# 5 years 2000-2004
unique(leaf_disc$daysout_number)
# lots of different times for the number of days left out, will have to standardize that
```

Catchments 31, 33, and 34 are the treatment streams. 

31 - high intensity harvest (89% removed)
33 - low intensity harvest (29% removed)
34U - medium intensity (42% removal)
34L - high intensity (89%)
34M - upstream of 34U and 34L

## Summarise
Because there are different time legths that they were left out for, I will need to standardize by the number of days left out. I also want to average the mass loss by year because they all have different numbers of sampling efforts.
```{r}
leaf_disc_sum<-leaf_disc %>% mutate(massloss_g_perday=massloss_g/daysout_number) %>% group_by(catchment, year) %>% summarise(mean_massloss=mean(massloss_g_perday))
```

## Plot
Now I can plot the mass loss by catchment.
```{r}
leaf_disc_sum$year<-as.character.Date(leaf_disc_sum$year)

ggplot(leaf_disc_sum)+
  geom_col(aes(x=catchment, y=mean_massloss, fill=year))
```

Not really any patterns for the catchments that were logged

```{r}
leaf_disc_sum2<-leaf_disc %>% mutate(massloss_g_perday=massloss_g/daysout_number) %>% group_by(catchment, year, position) %>% summarise(mean_massloss=mean(massloss_g_perday))
```

Now I can plot the mass loss by catchment.
```{r}
leaf_disc_sum2$year<-as.character.Date(leaf_disc_sum2$year)

ggplot(leaf_disc_sum2)+
  geom_col(aes(x=position, y=mean_massloss, fill=catchment), position="dodge")
```


# Leaf pack decomposition

## Load data
```{r}
leaf_pack<-read.csv("Turkey lake/TLW_leafpackDecomposition.csv")
head(leaf_pack)
str(leaf_pack)
```

Looking at the number of unique catchments, years, etc.
```{r}
unique(leaf_pack$catchment)
# only 2 catchments, both are unlogged
unique(leaf_pack$year)
# 2 years - 2003-2004
unique(leaf_pack$daysout_number)
# lots of different times for the number of days left out, will have to standardize that
```

Density plot of leaf pack
```{r}
ggplot(leaf_pack)+
  geom_density(aes(x=massloss_g))
```

Kind of a weird distribution but okay.

Need to standardize by the number of days left out and I will also average by year.
```{r}
leaf_pack_sum<-leaf_pack %>% mutate(massloss_g_perday=massloss_g/daysout_number) %>% group_by(catchment, year) %>% summarise(mean_massloss=mean(massloss_g_perday))
```

Now I can plot the mass loss by catchment.
```{r}
leaf_pack_sum$year<-as.character.Date(leaf_pack_sum$year)

ggplot(leaf_pack_sum)+
  geom_col(aes(x=catchment, y=mean_massloss, fill=year))
```

Alright, looks pretty not interesting. But maybe I am interested in the difference between buried and not buried?

```{r}
leaf_pack_sum2<-leaf_pack %>% mutate(massloss_g_perday=massloss_g/daysout_number) %>% group_by(catchment, year, position) %>% summarise(mean_massloss=mean(massloss_g_perday))
```

Now I can plot the mass loss by catchment.
```{r}
leaf_pack_sum2$year<-as.character.Date(leaf_pack_sum2$year)

ggplot(leaf_pack_sum2)+
  geom_col(aes(x=position, y=mean_massloss, fill=year), position="dodge")
```

It looks like there is slightly less mass loss when the disc is buried.

I am going to average over the years but plot by catchment.

```{r}
leaf_pack_sum3<-leaf_pack %>% mutate(massloss_g_perday=massloss_g/daysout_number) %>% group_by(catchment, position) %>% summarise(mean_massloss=mean(massloss_g_perday))
```

Now I can plot the mass loss by catchment.
```{r}
ggplot(leaf_pack_sum3)+
  geom_col(aes(x=position, y=mean_massloss, fill=catchment), position="dodge")
```

Alright, this really isn't interesting.

# Benthic invertebrate data
This is the invertebrate data from Turkey lakes. It has density of each invertebrate for each stream. It is currently in wide form, with each column as an invertebrate.
```{r}
inverts<-read.csv("TLW_invertebrateDensity.csv")
head(inverts)
```

I want to see the number of unique years and catchments.
```{r}
unique(inverts$catchment)
# 11 catchments, with 33 (low intensity), 34U (medium intensity), and 34L (high intensity) as the treatment streams
unique(inverts$year)
# 9 years, from 1995-2009 (not all years in between)
```

## Species abundance curves
We talked in class about creating rank abundance curves for species. Right now they are in "density" so I will start by converting all of them to counts by multiplying by 0.33. I will also put it into long form so that there is a column for invertebrate species. 

Turning it into long format

IGORS CODE:
```{r}
# Tidy data for plotting
inverts_long<-pivot_longer(inverts, "Aeshna":"Trichoptera", names_to = "species", values_to = "density") %>%
    mutate(count = density * 0.33)
```

NEED TO MAKE SURE REPLICATE IS INCLUDED HERE!!!
```{r}
# inverts_long2<-pivot_longer(inverts, cols=5:97, names_to="species", values_to="density")
```

Converting count to density. Also turning all of the NA values into 0s for both density and count
```{r}
inverts_long$count[is.na(inverts_long$count)] <- 0
```

I also want to add a column for the treatments.
```{r}
# To do
```

First I will just do a really simple plot of all of the count data added up over all of the years. I will take out all 0s because there are too many species and those are not interesting.

Making a data table with total values and filtering by the top 20 species so the plot doesnt get too crazy
```{r}
invert_sum<-inverts_long %>% group_by(species) %>% summarise(total_count=sum(count)) %>% 
  top_n(n=40, wt=total_count)
```

Plot the top species
```{r}
ggplot(invert_sum, mapping=aes(y = reorder(species, total_count), x=total_count))+
  geom_col()+
  labs(x="Total Species Count", y="Invertebrate Species")+
  theme_classic()
```

Can also facet by year

```{r}
invert_sum_yr<-inverts_long %>% group_by(species, year) %>% summarise(total_count=sum(count)) %>% 
  top_n(n=10, wt=total_count)
```


```{r}
ggplot(invert_sum_yr, mapping=aes(y = reorder(species, total_count), x=total_count))+
  geom_col()+
  labs(x="Total Species Count", y="Invertebrate Species")+
  facet_wrap(~year)+
  theme_classic()
```

This is bad lol.


## Codyn functions

```{r}
library(codyn)
head(inverts_long)
```
Need to make a date column and merge year and month togther into one column (k this is not working im annoyed)
```{r}
inverts_long$year<-as.numeric(inverts_long$year)
inverts_long$date <- zoo::as.yearmon(paste(inverts_long$year, inverts_long$month, sep="-"), "%Y-%M")
```

I am instead going to pool data by year.
```{r}
head(inverts_long)

unique(inverts_long$catchment)

catchment33 <- inverts_long %>%
  filter(catchment == "33", year == c(1995, 2001)) %>%
  group_by(year, replicate, species) %>%
  drop_na(count) %>% 
  summarise(count = sum(count))

```


Trying the abundance change function, which calculates diversity change in 2 time points.
```{r}
abundance_change(df=catchment33, time.var="year", species.var="species", abundance.var="count", replicate.var="replicate")
```

This calculates species turnover between two time points.
```{r}
turnover(df=catchment33, time.var="year", species.var="species", abundance.var="count", replicate.var="replicate")
```

Now I want to look at differences between the treatment and reference streams.

```{r}
unique(inverts_long$catchment)

catchments <- inverts_long %>%
  filter(catchment == c("34L", "42"), year == c(1998)) %>%
  group_by(year, replicate, species) %>%
  drop_na(count) %>% 
  summarise(count = sum(count))
```

Now I will try to compare between them. This is technically meant for paired samples but trying it.
```{r}
abundance_difference(df=catchments, time.var="year", species.var="species", abundance.var="count", replicate.var="replicate")
```










